Source Separation Methods for Computer-Assisted Orchestration

### 2022-04-20 source_separation/paper/paper.tex
Paper with AIMC template

% LaTeX template for MLSP papers. To be used with:
%   * mlspconf.sty - ICASSP/ICIP LaTeX style file adapted for MLSP, and
%   * IEEEbib.bst - IEEE bibliography style file.
% --------------------------------------------------------------------------
\documentclass{article}
\usepackage{amsmath,graphicx,mlspconf}
%%%%%%%%%%%%%%%%%%%%%%%% Some useful packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%% See related documentation %%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[caption=false, font=footnotesize]{subfig}% Modern replacement for subfigure package
\usepackage{paralist}% extended list environments
\usepackage{url}
\usepackage{hyperref}
\usepackage[figure,table]{hypcap}% hyperref companion
\usepackage{multirow}
\usepackage{amssymb}

\renewcommand{\baselinestretch}{0.96} % use the smallest value to make our paper 6 pages long (including references!)

% Copyright notices.
% ------------------
% Select one of the four copyright notices below. Only required for the camera-ready paper submission.
% 
% * For papers in which all authors are employed by the US government:
%\copyrightnotice{U.S.\ Government work not protected by U.S.\ copyright}

% * For papers in which all authors are employed by a Crown government (UK, Canada, and Australia):
%\copyrightnotice{978-1-7281-6338-3/21/\$31.00 {\copyright}2021 Crown}

% * For papers in which all authors are employed by the European Union:
%\copyrightnotice{978-1-7281-6338-3/21/\$31.00
%{\copyright}2021 European Union}

% * For all other papers:
\copyrightnotice{978-1-7281-6338-3/21/\$31.00 {\copyright}2021 IEEE}

% Header
\toappear{2021 IEEE International Workshop on Machine Learning for Signal Processing, Oct.\ 25--28, 2021, Gold Coast, Australia}

% Example definitions.
% --------------------
%\def\x{{\mathbf x}}
%\def\L{{\cal L}}

% Title.
% ------
\title{SOURCE SEPARATION METHODS FOR COMPUTER-ASSISTED ORCHESTRATION}
%
% Double-blind peer review.
% -------------------------
% Anonymize your paper for the double-blind peer-review process using the 
% following author and affiliation.
\name{Anonymous\thanks{Anonymous.}}
\address{Anonymous}

% Single address.
% ---------------
%\name{Author(s) Name(s)\thanks{Thanks to XYZ agency for funding.}}
%\address{Author Affiliation(s)}

% For example:
% ------------
%\address{%
%    School \\
%    Department \\
%    Address
%}
%
% Two addresses.
% --------------
%\twoauthors{%
%    A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}
%}{%
%    School A-B \\
%    Department A-B \\
%    Address A-B \\
%    Email A-B
%}{%
%   C. Author-three, D. Author-four\sthanks{The fourth author performed the work while at ...}
%}{%
%    School C-D \\
%    Department C-D \\
%    Address C-D \\
%    Email C-D
%}
% 
% Two or more addresses (alternative form).
% -----------------------------------------
% If you need to list more than 2 authors or the option for two options above 
% produces a poor author block, please use the following structure:
%\name{%
%    Author Name$^{\star \dagger}$%
%    \qquad Author Name$^{\star}$%
%    \qquad Author Name$^{\dagger}$\thanks{Thanks to XYZ agency for funding.}%
%}
%\address{%
%    $^{\star}$ Affiliation Number One \\%
%    $^{\dagger}$ Affiliation Number Two%
%}

\begin{document}
%\ninept

@@

\begin{abstract}
  In this paper, we will study the possibility of adding source separation as a pre-processing step to the computer-assisted orchestration process. We first discuss the motivation of this addition and its potential to increase the quality of orchestrations of multi-layered sounds. Second, we select several state-of-the-art models for both music source separation (separation of instruments) and universal sound separation (separation of arbitrary sounds of different types), and compare their effectiveness for the task of orchestration. We assess which methods best suit the needs of orchestration by applying them on hand picked target sounds, orchestrating the separated outputs, and finally comparing them to the orchestration of the same target without separation. Our experiments show that the quality of orchestrations improves, both qualitatively and quantitatively, indicating that our approach is promising. Finally, we compare unsupervised methods to supervised methods for separation, and comment on the effect of training data selection on performance of supervised methods.
\end{abstract}
%
\begin{keywords}
computer-assisted orchestration, sound source separation, Orchidea
\end{keywords}
%
\section{Introduction}\label{sec:introduction}
  Computer-assisted composition is a field that focuses on the creation of computational tools to aid in the musical composition process \cite{FerVic2013, Ari2005}. Within this field, target-based computer-assisted musical orchestration is an example of how musical creativity can be supported by tools such as machine learning.

	Orchestration~\cite{adler1989study} is hard to define, however one could say that orchestration is the set of musical writing techniques that focus on timbre, or \textit{tone color} of the sound. A simplistic  definition of musical orchestration would be to say that it consists of the distribution of the different musical voices of a piece to the instruments of the orchestra. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes.

	 Among the different approaches and techniques of computer-assisted orchestration, one of them, \textit{target-based orchestration}~\cite{Maresz2003}. Target-based computer-assisted orchestration helps composers explore new timbral possibilities by combining instrumental samples in such a way as to mimic the timbre of a given target sound. This target corresponds to an audio extract, of the composer's choice, which can be musical or not. Formally, it is the process of creating an orchestral score that best matches an arbitrary target sound given a similarity metric and constraints.

  Static orchestration is a type of orchestration that considers the target as a single vector of timbral features and creates only a single onset of note(s) as the solution, no matter if the target contains multiple onsets of sounds. In our experiments, we only perform static orchestrations.

@@

  We think source separation could be an important addition to improve orchestration as it allows for the orchestration of more complex sounds in a way that makes sense in an orchestral context. Orchestral music often uses multiple layers, so by separating the layers of a target and individually orchestrating them, the resulting orchestration may improve.

@@

%  The codebase for this paper can be found at: \url{https://github.com/dzluke/MLSP2021}, and you can listen to a selection of targets and orchestrations here: \url{https://dzluke.github.io/MLSP2021/}.
  The codebase for this paper can be found at: \url{anonymous}, and a selection of targets and orchestrations are available for listenig at: \url{anonymous}.

@@

  We compare the effectiveness of different source separation methods for the task of orchestration by applying the separation methods to various targets, orchestrating the separated output of the method, and finally comparing this to the orchestration of the target if no separation was performed (see Fig. \ref{fig:full_diagram}). Following the previous example, consider a target sound that is a combination of a low droning sound and high-pitched whistle. With separation, these two sounds would be disentangled into two separate sub-targets and each would be individually orchestrated. The separated orchestrations would then be combined to create the solution. This would then be compared to the orchestration when no separation had been performed.

  \begin{figure*}[t]
%  \begin{figure}[t]
    \centering
      \includegraphics[width=0.7\textwidth]{figures/diagram.png}
      \caption{Diagram comparing the process of orchestration with and without separation. At the top, the full target orchestration is created simply by orchestrating the target. At the bottom, the separation process is performed, and the resulting sub-targets are individually orchestrated with subsets of the orchestra before being recombined to obtain the full orchestration.}\label{fig:full_diagram}
%  \end{figure}  
  \end{figure*}

@@

    \subsection{Testing}\label{subsec:testing}
    In order to compare the effectiveness of the separation methods, we compare them with \textit{a full target orchestration}. The \textbf{full target orchestration} is the orchestration of the target without any separation performed.

    The procedure we use for testing is as follows:
    \begin{enumerate}
      \item The target is created as a combination of four randomly chosen sources, which are offset to begin playing at different times.
      \item The target, without any separation performed, is orchestrated using the entire orchestra. This is called the full target orchestration.
      \item Each separation method is performed on the target, each splitting the target into four sub-targets. 
      \item The full orchestra is randomly split into four equal sized subsets, each containing 7 to 8 instruments.
      \item For each method, the four sub-targets are separately orchestrated, each using a different subset of the orchestra, and then each orchestration is combined to play simultaneously. This creates the \textbf{separated orchestrations}.
      \item The distances between target and orchestration are calculated for the full target orchestration and the separated orchestration.
    \end{enumerate}

@@

    \begin{equation}\label{eqn:distance}
d(x, \tilde{x}) =\lambda_1 \sum_k \delta_{k1}(x_k - \tilde{x}_k) + \lambda_2 \sum_k \delta_{k2}|x_k - \tilde{x  }_k| \\
\end{equation}
where $\delta_{k1} = 1 \text{  if  } x_k \ge \tilde{x}_k, 0 \text{  otherwise}$; and $\delta_{k2} = 1 \text{  if  } x_k < \tilde{x}_k, 0 \text{  otherwise}$.

@@

    \subsection{Results}
    The average distances for full target and separated orchestrations are displayed in Table \ref{tab:distances}. These results are the average distance between target and orchestration for 300 different targets. We found that NMF performed the best out of all separation methods. TDCN++, Open-Unmix, and NMF performed better than the full target orchestrations, meaning that separating targets using these methods leads to a better orchestration than if no separation is applied. However, Demucs and TDCN performed worse.

  \section{Discussion}\label{sec:discussion}
  In this paper, we addressed the problem of using source separation as a pre-processing step for computer-assisted orchestration. The results show that performing source separation followed by a static orchestration of each individual sub-target improved the spectral distance between the original target and its orchestrated version in three out of five of the methods tested. We believe this shows that source separation is a useful pre-processing stage for computer-assisted orchestration. We also conducted a qualitative assessment of the orchestrations by acoustic inspection and found that the separated orchestrations are better from a musical standpoint.

  Furthermore, the one unsupervised method we tested, NMF, outperformed all the supervised methods. This suggests that methods like NMF can still be useful for specific problems when available data is limited. It is logical that unsupervised methods will perform well for problems in which there is limited data, as they require less data than supervised methods. It is important to notice that for the supervised methods, we used pre-trained versions of the neural models. All of these models, except for TDCN++, were trained on musical data. However, we tested them on a wide range of targets, many of which were not strictly musical. This could explain why Demucs, Open-Unmix, and TDCN performed worse than NMF, and why TDCN++, which was trained on arbitrary sounds, outperformed the other supervised methods. Therefore, we may infer from our results that models trained on generic sounds generate a better separation for orchestration when compared to models trained only on musical sounds.

  Since unsupervised methods perform well, it is reasonable to think that supervised methods trained on an appropriate dataset of arbitrary sounds could yield better overall performance for orchestration.

@@

% References should be produced using the bibtex program from suitable
% BiBTeX files (here: strings, refs, manuals). The IEEEbib.bst bibliography
% style file from IEEE produces unsorted bibliography list.
% -------------------------------------------------------------------------
\bibliographystyle{IEEEbib}
\bibliography{refs}

\end{document}

>>>>

% License: LaTeX Project Public License 1.3c
% file aimc2022.tex
% This is the LaTeX source for the instructions to authors using
% the LaTeX2e package 'aimc22.sty' for contributions to
% the Conference on AI Music Creativity.

\documentclass[letterpaper]{article}  %TODO: make sure this is the right format

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading aimc2022

% load package
\usepackage{aimc2022}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}       % figures

\bibliographystyle{unsrtnat}

\title{Source Separation Methods for Computer-Assisted Orchestration}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
  Single Author\thanks{Use footnote for providing further information
    about author (webpage, alternative address)---\emph{not} for acknowledging
    funding agencies.} \\
  Department of Music\\
  Strawberry-Raspberry University\\
  Pittsburgh, PA 15213 \\
  \texttt{rhino@music.strawberry-raspberry.edu} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}

@@

\begin{abstract}
  In this paper, we will study the possibility of adding source separation as a pre-processing step to the computer-assisted orchestration process. We first discuss the motivation of this addition and its potential to increase the quality of orchestrations of multi-layered sounds. Second, we select several state-of-the-art models for both music source separation (separation of instruments) and universal sound separation (separation of arbitrary sounds of different types), and compare their effectiveness for the task of orchestration. We assess which methods best suit the needs of orchestration by applying them on hand picked target sounds, orchestrating the separated outputs, and finally comparing them to the orchestration of the same target without separation. Our experiments show that the quality of orchestrations improves, both qualitatively and quantitatively, indicating that our approach is promising. Finally, we compare unsupervised methods to supervised methods for separation, and comment on the effect of training data selection on performance of supervised methods.
\end{abstract}

\section{Introduction}\label{sec:introduction}
  Computer-assisted composition is a field that focuses on the creation of computational tools to aid in the musical composition process \cite{FerVic2013, Ari2005}. Within this field, target-based computer-assisted musical orchestration is an example of how musical creativity can be supported by tools such as machine learning. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes. In particular, target-based computer-assisted orchestration helps composers explore new timbral possibilities by combining instrumental samples in such a way as to mimic the timbre of a given target sound. Formally, it is the process of creating an orchestral score that best matches an arbitrary target sound given a similarity metric and constraints~\cite{Maresz2003}. Static orchestration is a type of orchestration that considers the target as a single vector of timbral features and creates only a single onset of note(s) as the solution, no matter if the target contains multiple onsets of sounds. In our experiments, we only perform static orchestrations.

@@

%  Source separation is an important addition to improve orchestration as it allows for the orchestration of more complex sounds in a way that makes sense in an orchestral context. Orchestral music often uses multiple layers, so by separating the layers of a target and individually orchestrating them, the resulting orchestration may improve.

@@

%  The codebase for this paper can be found at: \url{https://github.com/dzluke/MLSP2021}, and you can listen to a selection of targets and orchestrations here: \url{https://dzluke.github.io/MLSP2021/}.
  The codebase for this paper can be found at: \url{anonymous}, and you can listen to a selection of targets and orchestrations here: \url{anonymous}.

@@

  We compare the effectiveness of different source separation methods for the task of orchestration by applying the separation methods to various targets, orchestrating the separated output of the method, and finally comparing this to the orchestration of the target if no separation was performed (see Fig. \ref{fig:full_diagram}). Following the previous example, consider a target sound that is a combination of a low droning sound and high-pitched whistle. With separation, these two sounds would be disentangled into two separate sub-targets and each would be individually orchestrated. The separated orchestrations would then be combined to create the solution. This would then be compared to the orchestration if no separation had been performed.

  \begin{figure}[t]
    \centering
      \includegraphics[width=\columnwidth]{figures/diagram.png}
      \caption{Diagram comparing the process of orchestration with and without separation. At the top, the full target orchestration is created simply by orchestrating the target. At the bottom, the separation process is performed, and the resulting sub-targets are individually orchestrated with subsets of the orchestra before being recombined to obtain the full orchestration.}\label{fig:full_diagram}
  \end{figure}

@@

    \subsection{Testing}\label{subsec:testing}
    In order to compare the effectiveness of the separation methods, a full target orchestration and a ground truth orchestration are done for each target. The \textbf{full target orchestration} is the orchestration of the target without any separation performed. The \textbf{ground truth orchestration} takes the four sources that make up the target and orchestrates each one separately, then combines the orchestrations. This ground truth represents the orchestration that could take place if the separation method was perfect and separated a target exactly into its constituent sources.

    The procedure we use for testing is as follows:
    \begin{enumerate}
      \item The target is created as a combination of four randomly chosen sources, which are offset to begin playing at different times.
      \item The target, without any separation performed, is orchestrated using the entire orchestra. This is called the full target orchestration.
      \item Each separation method is performed on the target, each splitting the target into four sub-targets.
      \item The full orchestra is randomly split into four equal sized subsets, each containing 7 to 8 instruments.
      \item For each method, the four sub-targets are separately orchestrated, each using a different subset of the orchestra, and then each orchestration is combined to play simultaneously. This creates the \textbf{separated orchestrations}.
      \item The ground truth orchestration is created by orchestrating the four sources, using the same orchestra subsets from step 4, and then combined to play simultaneously.
      \item The distances between target and orchestration are calculated for the full target orchestration, the separated orchestration, and the ground truth orchestration.
    \end{enumerate}

@@

    \begin{equation}\label{eqn:distance}
d(x, \tilde{x}) =\lambda_1 \sum_k \delta_{k1}(x_k - \tilde{x}_k) + \lambda_2 \sum_k \delta_{k2}|x_k - \tilde{x  }_k| \\
\end{equation}
where $\delta_{k1} = 1 \textrm{  if  } x_k \ge \tilde{x}_k, 0 \textrm{  otherwise}$; and $\delta_{k2} = 1 \textrm{  if  } x_k < \tilde{x}_k, 0 \textrm{  otherwise}$.

@@

    \subsection{Results}
    The average distances for full target, separated, and ground truth orchestrations are displayed in Table \ref{tab:distances}. These results are the average distance between target and orchestration for 300 different targets. We found that NMF performed the best out of all separation methods, and even outperformed the ground truth orchestrations, on average. TDCN++, Open-Unmix, and NMF performed better than the full target orchestrations, meaning that separating targets using these methods leads to a better orchestration than if no separation is applied.

  \section{Discussion}\label{sec:discussion}
  In this paper, we addressed the problem of using source separation as a pre-processing step for computer-assisted orchestration. The results show that performing source separation followed by a static orchestration of each individual sub-target improved the spectral distance between the original target and its orchestrated version in three out of five of the methods tested. This shows that source separation is a useful pre-processing stage for computer-assisted orchestration.

  Furthermore, the one unsupervised method we tested, NMF, outperformed all the supervised methods. This suggests that methods like NMF can still be useful for specific problems when available data is limited. It is logical that unsupervised methods will perform well for problems in which there is limited data, as they require less data than supervised methods. NMF also outperformed the ground truth orchestrations, which could be due to the relative simplicity of NMF's output subtargets making subsequent orchestration easier for Orchidea, but more experimentation would need to be done in order to confirm this. It is important to notice that for the supervised methods, we used pre-trained versions of the neural models. All of these models, except for TDCN++, were trained on musical data. However, we tested them on a wide range of targets, many of which were not strictly musical. This could explain why Demucs, Open-Unmix, and TDCN performed worse than NMF, and why TDCN++, which was trained on arbitrary sounds, outperformed the other supervised methods. Therefore, we may infer from our results that models trained on generic sounds generate a better separation for orchestration when compared to models trained only on musical sounds.

  Since unsupervised methods perform well, it is reasonable to think that supervised methods trained on a dataset of arbitrary sounds could yield better overall performance for orchestration.

@@

\bibliography{bibliography}
\end{document}
